Assessing exposure to Atlantic Basin tropical storms in United States counties

### 2016-08-03 writing/DraftExposurePaper/DraftExposurePaper.Rmd
Work on draft

```{r echo = FALSE, message = FALSE, warning = FALSE}
library(hurricaneexposuredata)
library(hurricaneexposure)
library(knitr)
library(ggplot2)
library(ggthemes)
library(grid)
library(gridExtra)
library(plyr)
library(dplyr)
library(tidyr)
library(choroplethr)
library(RColorBrewer)
library(weathermetrics)

#devtools::install_github("ropenscilabs/rnoaa")
library(rnoaa)
#devtools::install_github("leighseverson/countyweather")
library(countyweather)

@@

Next, we looked at several different counties within a few notable storms. These storms were picked based on storms that had a large number of ground-based precipitation observations in either the main text or supplemental material of the *Monthly Weather Review* report.

[Move figures here for NLDAS vs. county average values.]

Finally, we compared exposure results based on the NLDAS-2 precipitation data with values from monitors reported for certain storms in the *Monthly Weather Reviews* yearly summary of the Atlantic basin tropical storm season (Figure \ref{fig:mwr_comparison}). Here are some of the plots comparing county-level estimates of rain exposure from the NLDAS-2 data to all *MWR*-reported rainfall estimates within the county:

@@

load("frances_gauge.RData")
frances_gauge$rain_mm <- convert_precip(frances_gauge$precip_in,
                                        "inches", "mm")
frances_mwr <- plot_mwr_compare(frances_gauge)

load("ike_gauge.RData")
ike_gauge$rain_mm <- convert_precip(ike_gauge$rain_in, 
                                    "inches", "mm")
ike_mwr <- plot_mwr_compare(ike_gauge)

load("alex_gauge.Rdata")
alex_mwr <- plot_mwr_compare(alex_gauge)

@@

```{r}
#Find the top 15 FIPS by number of reporting stations during Ike-2008 (Warning: Takes a long time)
#my_fips <- ike_fips[c(1:6, 10, 13:16, 18, 20:25, 28:29, 31, 33:34, 36:40, 42, 44:45)]
#my_stations <- NA
#ike_table <- expand.grid(my_fips, my_stations)
#colnames(ike_table) <- c("fips", "station")
#for(fips in my_fips){
        #ike_weather <- weather_fips(fips = fips,
                                    #date_min = "2008-09-01",
                                    #date_max = "2008-09-14",
                                    #var = "PRCP")
        #ike_table[ike_table$fips == fips, "station"] <- max(ike_weather$daily_weather$prcp_reporting)    
#} 
#Ike FIPS without data: 7, 8, 9, 11, 12, 17, 19, 26, 27, 30, 32,35, 41, 43, 
#ike_table <- ike_table %>%
        #arrange(desc(station)) %>%
        #slice(1:15)
#ike_top <- ike_table$fips
#Find the top 15 FIPS by number of reporting stations for Alex-2010:
#my_fips <- alex_fips[c(1:2, 4:6, 8, 10, 12:14, 16:18)]
#my_stations <- NA
#alex_table <- expand.grid(my_fips, my_stations)
#colnames(alex_table) <- c("fips", "station")
#for(fips in my_fips){
#        alex_weather <- weather_fips(fips = fips,
#                                    date_min = "2008-09-01",
#                                    date_max = "2008-09-14",
#                                    var = "PRCP")
#        alex_table[alex_table$fips == fips, "station"] <- max(alex_weather$daily_weather$prcp_reporting)    
#} 
#Alex FIPS without data: 3, 7, 9, 11, 15, 19   
#alex_table <- alex_table %>%
#        arrange(desc(station)) %>%
#        slice(1:15)
#alex_top <- alex_table$fips
#fips_table <- rbind(alex_table, ike_table)
#fips_table <- fips_table %>%
#        arrange(desc(station)) %>%
#        slice(1:15)
#all_fips <- fips_table$fips
#all_fips <- unique(all_fips)
```

```{r eval = FALSE}
county_timeseries(fips = ike_fips[1:2], 
                  percent_coverage = 0, 
                  date_min = "2008-09-01", 
                  date_max = "2010-09-14", var = "PRCP", 
                  out_directory = "ike_ex")
ike_1 <- readRDS("ike_ex/12021.rds")
keys_cd <- subset(closest_dist, closest_dist$storm_id == "Ike-2008" & closest_dist$fips == "12087")
keys_cd <- keys_cd$closest_date
ike_1 <- ike_1 %>%
        select(date, prcp) %>%
        filter(date %in% seq(from = as.Date("2008-09-07"), to = as.Date("2008-09-10"), by = 1))
ike_rain <- subset(rain, rain$storm_id == "Ike-2008" & rain$fips == "12087")
ike_sum <- sum(ike_1$prcp)
ike_nldas_sum <- subset(ike_nldas, ike_nldas$fips == "12087")
ike_nldas_sum <- ike_nldas_sum$tot_precip
```

```{r eval = FALSE}
options("noaakey" = Sys.getenv("noaakey"))
#Alex-2010
alex_fips <- alex_nldas$fips
alex_fips <- unique(alex_fips)

#Ike-2008
ike_fips <- ike_nldas$fips
ike_fips <- unique(ike_fips)

#Allison-2010
allison_fips <- allison_nldas$fips
allison_fips <- unique(allison_fips)

#Frances-2004
frances_fips <- frances_nldas$fips
frances_fips <- unique(frances_fips)
```

Trying to create a loop to identify the FIPS codes with the greatest number of stations reporting for each time period.

```{r eval = FALSE}
my_fips <- ike_fips[1:3]
my_stations <- NA
test <- expand.grid(my_fips, my_stations)
colnames(test) <- c("fips", "station")
for(fips in my_fips){
        ike_weather <- weather_fips(fips = fips,
                                    date_min = "2008-09-01",
                                    date_max = "2008-09-14",
                                    var = "PRCP")
        test[test$fips == fips, "station"] <- max(ike_weather$daily_weather$prcp_reporting)    
} #This works fine (small subset of code below)
ike_table <- expand.grid(ike_fips, my_stations)
colnames(ike_table) <- c("fips", "station")
#for(fips in ike_fips){
        ike_weather <- weather_fips(fips = fips,
                                    date_min = "2008-09-01",
                                    date_max = "2008-09-14",
                                    var = "PRCP")
        ike_table[ike_table$fips == fips, "station"] <- max(ike_weather$daily_weather$prcp_reporting)
} #getting error
```

Our exposure metrics restricts "exposed" counties to be within 500 km of the storm track. In a few cases, hurricane best tracks stop when the storm transitions from a tropical to extratropical system [let's check that I got the wording right on this]. In these cases, extreme rain from the storm system might have affected counties further along the storm's path, but that are not counted as being near the storm's track. For example, Lee in 2011 was only tracked through the best tracks until it was near the Georgia-Tennessee border. However, remnants of the system brought extreme rain further north, into the Mid-Atlantic and New England [I believe that the Monthly Weather Review for the 2011 season includes a description of these rain impacts. We likely want to look over that and incorporate it here]:

```{r echo = FALSE, fig.width = 6, fig.height = 8}
storm <- "Lee-2011"
a <- map_tracks(storms = storm, plot_points = FALSE, 
                plot_object = map_counties(storm = storm, 
                                           metric = "rainfall")) + 
        ggtitle("Rainfall during Lee, 2011")
b <- map_tracks(storms = storm, plot_points = FALSE, 
                plot_object = map_rain_exposure(storm = storm, 
                                           dist_limit = 500,
                                           rain_limit = 75)) + 
        ggtitle("Rain-based exposure during Lee, 2011")
grid.arrange(a, b, ncol = 1)
```

Other possible examples of this during our study period include Nicole (2010) [check storm history in MWR to confirm that Mid-Atlantic and New England rain during this time was related to storm], Ida (2009), Humberto (2007), Katrina (2005), Grace (2003) [confirm rain was from storm remnants], Frances (1998), and Marco (1990).

For rain-based exposure, we use a distance constraint of 500 km. With storms where the best track extends throughout the storm, this should typically be a large enough swath to include all extreme rain exposures from the storm. There are only a few very large storms where this might not be the case. One example is Ike (2008), in which case the 500 km constraint excludes some of the counties in northern Texas that got heavy rain from the system:

```{r echo = FALSE, fig.width = 6, fig.height = 8}
storm <- "Ike-2008"
a <- map_tracks(storms = storm, plot_points = FALSE, 
                plot_object = map_counties(storm = storm, 
                                           metric = "rainfall")) + 
        ggtitle("Rainfall during Ike, 2008")
b <- map_tracks(storms = storm, plot_points = FALSE, 
                plot_object = map_rain_exposure(storm = storm, 
                                           dist_limit = 500,
                                           rain_limit = 75)) + 
        ggtitle("Rain-based exposure during Ike, 2008")
grid.arrange(a, b, ncol = 1)
```

Other large systems where this might be a concern include Wilma (2005), Hermine (2004) [confirm rain in SC, NC, and VA during this time was from this storm], Arlene (1993) [confirm rain through to LA is from this storm].

@@

```{r}
#Ike-2008 wind
load("ike_wind.RData")
ike_wind <- ike_wind %>%
        filter(!is.na(max_sust_kt)) %>%
        mutate(max_sust_mps = convert_wind_speed(max_sust_kt, 
                                                 old_metric = "knots",
                                                 new_metric = "mps"),
               fips = sprintf("%05s", fips))

@@

```{r eval = FALSE}
#frances_wind <- read.csv("/Users/joshuaferreri/Documents/NWR_frances_wind.csv", 
 #                        na.strings = 999,
  #                       header = TRUE)
#save(frances_wind, file = "writing/DraftExposurePaper/frances_wind.RData")
```

```{r}
load("frances_wind.RData")
frances_wind <- frances_wind %>%
        filter(!is.na(wind_sust)) %>%
        mutate(max_sust_mps = convert_wind_speed(wind_sust, 
                                                 old_metric = "knots",
                                                 new_metric = "mps"),
               fips = sprintf("%05s", fips))

@@

Map of the count of county exposures to the top 2,000 county-storm pairs by distance-based exposure, based on "exposed" = distance of 100 kilometers or less [GBA-- These are **awesome**. I think we want to tweak the color scale and the legend a bit so that they're consistent between the three plots showing the top 2,000 storm-county pairs for the three metrics (distance, rain, and wind). That will make it easier to compare across the three maps, which is what we want readers to be able to do here.]:

>>>>

```{r echo = FALSE, message = FALSE, warning = FALSE}
library(hurricaneexposuredata)
library(hurricaneexposure)
library(knitr)
library(ggplot2)
library(ggthemes)
library(grid)
library(gridExtra)
library(plyr)
library(dplyr)
library(tidyr)
library(choroplethr)
library(RColorBrewer)
library(weathermetrics)
library(lubridate)

#devtools::install_github("ropenscilabs/rnoaa")
library(rnoaa)
#devtools::install_github("leighseverson/countyweather")
options("noaakey" = Sys.getenv("noaakey"))
library(countyweather)

@@

Next, we looked at several different counties within a few notable storms, to compare county-averaged ground-based monitor values with our estimates (Figure \ref{fig:county_ave_comparison}). These storms were picked based on storms that had a large number of ground-based precipitation observations in either the main text or supplemental material of the *Monthly Weather Review* report.

```{r echo = FALSE, eval = FALSE}
options("noaakey" = Sys.getenv("noaakey"))

allison_gauge <- read.csv("MWR_gauge_rain_data/allison_mwr.csv") 
allison_fips <- sprintf("%05s", as.character(unique(allison_gauge$fips)))
allison_fips <- sample(allison_fips, 10)
date_range <- closest_dist %>% 
        filter(storm_id == "Allison-2001") %>%
        summarize(earliest_date = min(ymd(closest_date)) - ddays(2),
                  latest_date = max(ymd(closest_date)) + ddays(2))
county_timeseries(allison_fips, percent_coverage = 0,
                  date_min = format(date_range[1, 1], "%Y-%m-%d"),
                  date_max = format(date_range[1, 2], "%Y-%m-%d"),
                  var = "PRCP", out_directory = "county_ave/allison/")

load("MWR_gauge_rain_data/frances_gauge.RData")
frances_fips <- sprintf("%05s", as.character(unique(frances_gauge$fips)))
frances_fips <- sample(frances_fips, 10)
date_range <- closest_dist %>% 
        filter(storm_id == "Frances-2004") %>%
        summarize(earliest_date = min(ymd(closest_date)) - ddays(2),
                  latest_date = max(ymd(closest_date)) + ddays(2))
county_timeseries(frances_fips, percent_coverage = 0,
                  date_min = format(date_range[1, 1], "%Y-%m-%d"),
                  date_max = format(date_range[1, 2], "%Y-%m-%d"),
                  var = "PRCP", out_directory = "county_ave/frances/")

load("MWR_gauge_rain_data/ike_gauge.RData")
ike_fips <- sprintf("%05s", as.character(unique(ike_gauge$fips)))
ike_fips <- sample(ike_fips, 10)
date_range <- closest_dist %>% 
        filter(storm_id == "Ike-2008") %>%
        summarize(earliest_date = min(ymd(closest_date)) - ddays(2),
                  latest_date = max(ymd(closest_date)) + ddays(2))
county_timeseries(ike_fips, percent_coverage = 0,
                  date_min = format(date_range[1, 1], "%Y-%m-%d"),
                  date_max = format(date_range[1, 2], "%Y-%m-%d"),
                  var = "PRCP", out_directory = "county_ave/ike/")

load("MWR_gauge_rain_data/alex_gauge.RData")
alex_fips <- sprintf("%05s", as.character(unique(alex_gauge$fips)))
alex_fips <- sample(alex_fips, 10)
date_range <- closest_dist %>% 
        filter(storm_id == "Alex-2010") %>%
        summarize(earliest_date = min(ymd(closest_date)) - ddays(2),
                  latest_date = max(ymd(closest_date)) + ddays(2))
county_timeseries(alex_fips, percent_coverage = 0,
                  date_min = format(date_range[1, 1], "%Y-%m-%d"),
                  date_max = format(date_range[1, 2], "%Y-%m-%d"),
                  var = "PRCP", out_directory = "county_ave/alex/")
```

```{r county_ave_comparison, echo = FALSE, fig.width = 10, fig.height = 10, fig.align = "center", fig.cap = "Comparison of county-level average cumulative rainfall for the period from two days before to one day after the storm's closest approach to a county, using NLDAS-2 versus ground-based monitors, for counties with both estimates available for four example storms. The size of each point indicates the average number of monitors reporting from that county during the window (larger size indicates more monitors). The text over each point gives the FIPS county identifier."}
plot_county_ave_compare <- function(county_dir, storm_id){
        storm_title <- paste0(paste(unlist(strsplit(
                as.character(storm_id), "-")),
                             collapse = " ("), ")")
        storm_year <- gsub(".+-", "", storm_id)
        counties <- gsub(".rds", "",
                         list.files(paste0("county_ave/", county_dir)))

        storm_ave <- vector("list", length = length(counties))
        names(storm_ave) <- counties

        for(x in counties){
                county_weather <- readRDS(paste0("county_ave/", 
                                                 county_dir, "/", x,
                                                 ".rds"))$averaged %>%
                        dplyr::ungroup() %>%
                        dplyr::mutate(date = ymd(date)) %>%
                        dplyr::mutate(fips = x)
                storm_ave[[x]] <- county_weather
                }
        storm_ave <- do.call("rbind", storm_ave)

        storm_rain <- county_rain(counties = counties, 
                                  start_year = storm_year,
                                  end_year = storm_year,
                                  rain_limit = 0, dist_limit = 2000) %>%
                dplyr::filter(storm_id == storm_id)

        county_ave <- storm_ave %>%
                dplyr::left_join(storm_rain, by = "fips") %>%
                dplyr::group_by(fips) %>%
                dplyr::filter(ymd(closest_date) - ddays(2) <= date &
                       date <= ymd(closest_date) + ddays(1)) %>%
                dplyr::summarize(monitor_rain = sum(prcp),
                  tot_precip = first(tot_precip),
                  prcp_reporting = mean(prcp_reporting))
        plot_range <- range(county_ave[ , c("monitor_rain", "tot_precip")]) + 
                c(-15, 15)

        county_ave_plot <- ggplot(county_ave, 
                                  aes(x = tot_precip, 
                                      y = monitor_rain)) +
                geom_abline(aes(intercept = 0, slope = 1), color = "gray",
                    alpha = 0.5) +
                geom_point(aes(size = prcp_reporting), alpha = 0.3) +
                geom_text(aes(label = fips)) +
                theme_few() +
                scale_size_continuous(guide = "none") +
                ylab("Rainfall (mm) based on \naveraged county monitors") +
                xlab("Rainfall (mm) based on NLDAS-2 county data") +
                xlim(plot_range) +
                ylim(plot_range) + 
                ggtitle(storm_title)
        return(county_ave_plot)
}

allison_county_ave <- plot_county_ave_compare("allison", 
                                           storm_id = "Allison-2001")

frances_county_ave <- plot_county_ave_compare("frances", 
                                           storm_id = "Frances-2004")

ike_county_ave <- plot_county_ave_compare("ike", 
                                           storm_id = "Ike-2008")

alex_county_ave <- plot_county_ave_compare("alex", 
                                           storm_id = "Alex-2010")

grid.arrange(allison_county_ave, frances_county_ave,
             ike_county_ave, alex_county_ave, ncol = 2)
```

Finally, we compared exposure results based on the NLDAS-2 precipitation data with values from monitors reported for certain storms in the *Monthly Weather Reviews* yearly summary of the Atlantic basin tropical storm season (Figure \ref{fig:mwr_comparison}).

@@

load("MWR_gauge_rain_data/frances_gauge.RData")
frances_gauge$rain_mm <- convert_precip(frances_gauge$precip_in,
                                        "inches", "mm")
frances_mwr <- plot_mwr_compare(frances_gauge)

load("MWR_gauge_rain_data/ike_gauge.RData")
ike_gauge$rain_mm <- convert_precip(ike_gauge$rain_in, 
                                    "inches", "mm")
ike_mwr <- plot_mwr_compare(ike_gauge)

load("MWR_gauge_rain_data/alex_gauge.Rdata")
alex_mwr <- plot_mwr_compare(alex_gauge)

@@

Our exposure metrics restricts "exposed" counties to be within 500 km of the storm track. In a few cases, hurricane best tracks stop when the storm transitions from a tropical to extratropical system [let's check that I got the wording right on this]. In these cases, extreme rain from the storm system might have affected counties further along the storm's path, but that are not counted as being near the storm's track. For example, Lee in 2011 was only tracked through the best tracks until it was near the Georgia-Tennessee border. However, remnants of the system brought extreme rain further north, into the Mid-Atlantic and New England (Figure \ref{fig:lee_rain_example}). [I believe that the Monthly Weather Review for the 2011 season includes a description of these rain impacts. We likely want to look over that and incorporate it here]. Other possible examples of this during our study period include Nicole (2010) [check storm history in MWR to confirm that Mid-Atlantic and New England rain during this time was related to storm], Ida (2009), Humberto (2007), Katrina (2005), Grace (2003) [confirm rain was from storm remnants], Frances (1998), and Marco (1990).

```{r lee_rain_example, echo = FALSE, fig.width = 6, fig.height = 8, fig.align = "center", fig.cap = "Example of a tropical storm for which the 'best tracks' track ends early enough to miss significant later effects. These two maps show rainfall during the days surrounding Lee (2011) (top) as well as the counties classified as 'exposed' (bottom) based on criterion of (1) storm approaches within 500 kilometers of the county's population mean center and (2) county gets 75 millimeters or more of precipitation over the period between two days before and one day after the storm's closest approach to the county. The 'best track' is shown in red in both plots and illustrates that tracking ceased too early to capture more northern storm-associated rainfall."}
storm <- "Lee-2011"
a <- map_tracks(storms = storm, plot_points = FALSE, 
                plot_object = map_counties(storm = storm, 
                                           metric = "rainfall")) + 
        ggtitle("Rainfall during Lee, 2011")
b <- map_tracks(storms = storm, plot_points = FALSE, 
                plot_object = map_rain_exposure(storm = storm, 
                                           dist_limit = 500,
                                           rain_limit = 75)) + 
        ggtitle("Rain-based exposure during Lee, 2011")
grid.arrange(a, b, ncol = 1)
```

For rain-based exposure, we use a distance constraint of 500 km. With storms where the best track extends throughout the storm, this should typically be a large enough swath to include all extreme rain exposures from the storm. There are only a few very large storms where this might not be the case. One example is Ike (2008), in which case the 500 km constraint excludes some of the counties in northern Texas that got heavy rain from the system (Figure \ref{fig:ike_rain_example}). Other large systems where this might be a concern include Wilma (2005), Hermine (2004) [confirm rain in SC, NC, and VA during this time was from this storm], Arlene (1993) [confirm rain through to LA is from this storm].

```{r ike_rain_example, echo = FALSE, fig.width = 6, fig.height = 8, fig.align = "center", fig.cap = "Example of a storm large enough that some storm-associated rainfall was likely missed when using a 500 kilometer distance criterion with the rain-based exposure metric. These two maps show rainfall during the days surrounding Ike (2008) (top) as well as the counties classified as 'exposed' (bottom) based on criterion of (1) storm approaches within 500 kilometers of the county's population mean center and (2) county gets 75 millimeters or more of precipitation over the period between two days before and one day after the storm's closest approach to the county. The storm's 'best track' is plotted on each map in red. This example storm shows that some counties in northern Texas were incorrectly classified as 'unexposed' based on storm-associated rain because they fell outside the distance of 500 kilometers."}
storm <- "Ike-2008"
a <- map_tracks(storms = storm, plot_points = FALSE, 
                plot_object = map_counties(storm = storm, 
                                           metric = "rainfall")) + 
        ggtitle("Rainfall during Ike, 2008")
b <- map_tracks(storms = storm, plot_points = FALSE, 
                plot_object = map_rain_exposure(storm = storm, 
                                           dist_limit = 500,
                                           rain_limit = 75)) + 
        ggtitle("Rain-based exposure during Ike, 2008")
grid.arrange(a, b, ncol = 1)
```

@@

```{r}
#Ike-2008 wind
load("MWR_gauge_wind_data/ike_wind.RData")
ike_wind <- ike_wind %>%
        filter(!is.na(max_sust_kt)) %>%
        mutate(max_sust_mps = convert_wind_speed(max_sust_kt, 
                                                 old_metric = "knots",
                                                 new_metric = "mps"),
               fips = sprintf("%05s", fips))

@@

```{r}
load("MWR_gauge_wind_data/frances_wind.RData")
frances_wind <- frances_wind %>%
        filter(!is.na(wind_sust)) %>%
        mutate(max_sust_mps = convert_wind_speed(wind_sust, 
                                                 old_metric = "knots",
                                                 new_metric = "mps"),
               fips = sprintf("%05s", fips))

@@

Map of the count of county exposures to the top 2,000 county-storm pairs by distance-based exposure, based on "exposed" = distance of 100 kilometers or less:
